---
title: "Trait-dependent diversification"
author: "Rutger Vos (@rvosa)"
date: "12-12-2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load data

We are going to need a tree and a data set with information on grazing:

```{r load_data}
library(ape)
tree <- read.tree(file = "ungulates.nwk")
data <- read.table("ungulates-grazing.tsv", sep = "\t", header = T)

# remove species for which we don't have complete data
data <- data[!is.na(data$taxon_name),]
rownames(data) <- data$taxon_name

# remove species in the tree that are not in the data matrix
species.to.exclude <- tree$tip.label[!(tree$tip.label %in% data$taxon_name)]
tree <- drop.tip(tree,species.to.exclude)
data <- data[tree$tip.label,]

# clean memory
rm(species.to.exclude)
```

## Analyze diversification

Look at the diversification of the Ungulates:

```{r ltt}
library(phytools, quietly = T)
res<-ltt(tree)
```

Are there shifts in the diversification rate?

```{r shifts}
library(TreePar)

# make tree binary and ultrametric
binultra <- multi2di(force.ultrametric(tree, method = "extend"))

# assume a near complete tree, rho[1]=0.9
rho <- c(0.9,1)

# set windows of 10myr, starting 0, ending 90mya
grid <- 10
start <- 0
end <- 90

# estimate time, lambda, mu
x <- getx(binultra)
res <- bd.shifts.optim(x,c(rho,1),grid,start,end)[[2]]
```

Now what is the MLE of a single rate shift:

```{r MLE}
res[[2]][6]
```

In other words, 30mya. What is the rate before?

```{r rate_before}
res[[2]][5]
```

And after?

```{r rate_after}
res[[2]][4]
```

Significant?

```{r significant}
i<-1
test<-pchisq(2*(res[[i]][1]-res[[i+1]][1]),3)
test > 0.95
```

## BiSSE analysis

Now let's see if our approximation of grazing has anything to do with that rate shift.

```{r bisse}
library(diversitree)

# Convert the character into a numeric binary vector
char1 <- as.numeric(data$character_value)
names(char1) <- row.names(data)

# Create a BISSE likekihood model
lik <- make.bisse(binultra, char1)

# Create starting points for the search
p <- starting.point.bisse(binultra)

# Perform a ML search
fit <- find.mle(lik, p, method="subplex")

# lnL of the model
logLik(fit)

# The fitted parameters
round(coef(fit), 2)

# Test a constrained model where the speciation rates and extinction rates 
# are set equal
lik.l <- constrain(lik, lambda1 ~ lambda0, mu1 ~ mu0)
fit.l <- find.mle(lik.l, p[-c(1,3)], method="subplex")
logLik(fit.l)

# Fitted parameters
round(coef(fit.l), 2)

# Test for statistical difference between the two models
anova(fit, equal.lambda=fit.l)
```