---
title: "Imputing missing values"
author: "Rutger Vos (@rvosa)"
date: "8-12-2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Pre-processing data

We have multiple files with different data types, which we all load separately into data frames:

```{r load_dfs}
df_cont <- read.table("ungulates-continuous.tsv", sep = "\t", header = T, row.names = 1)
df_doms <- read.table("ungulates-domesticated.tsv", sep = "\t", header = T, row.names = 1)
```

Now we want to merge the data frames so that we get one big one, with all the columns as 
in the input frames, and the union of the rows across all frames:

```{r merge}
# SEE: https://adairama.wordpress.com/2017/11/22/how-to-merge-multiple-datasets-in-r-based-on-row-names/
# mimics a recursive merge or full outer join
multimerge <- function (mylist) {
  
  # get the union of the distinct taxon/row names
  unames <- unique(unlist(lapply(mylist, rownames)))
  n <- length(unames)

  # turns input df's into a matrix 'out' with all rows across mylist 
  out <- lapply(mylist, function(df) {
    tmp <- matrix(nr = n, nc = ncol(df), dimnames = list(unames,colnames(df)))
    tmp[rownames(df), ] <- as.matrix(df)
    rm(df); gc()
    return(tmp)
  })
 
  # check if all rownames match unames
  stopifnot( all( sapply(out, function(x) identical(rownames(x), unames)) ) )
  
  # cbind the matrices
  bigout <- do.call(cbind, out)
  
  # set row and column names
  colnames(bigout) <- paste(rep(names(mylist), sapply(mylist, ncol)), unlist(sapply(mylist, colnames)), sep = "_")
  return(bigout)
}

df <- data.frame( multimerge( list( cont=df_cont, doms=df_doms ) ) )
rm(df_cont,df_doms)
```

And now let's filter this result to remove all the columns that have more than 50% NA:

```{r filter_columns}
df <- df[, colMeans(is.na(df)) <= .5] 
```

And tidy up the remaining columns:

```{r col_factors}
# 0/1 factor
#df$doms_IsDomesticated <- as.factor(df$doms_IsDomesticated)

# factors with multiple, ordered levels
#df$disc_X1.1_ActivityCycle <- factor(x = df$disc_X1.1_ActivityCycle, ordered = T);
#df$disc_X12.1_HabitatBreadth <- factor(x = df$disc_X12.1_HabitatBreadth, ordered = T);
#df$disc_X12.2_Terrestriality <- factor(x = df$disc_X12.2_Terrestriality, ordered = T);
#df$disc_X6.1_DietBreadth <- factor(x = df$disc_X6.1_DietBreadth, ordered = T);
#df$disc_X6.2_TrophicLevel <- factor(x = df$disc_X6.2_TrophicLevel, ordered = T);

# absolute values for latitudes
df$cont_X26.2_GR_MaxLat_dd <- abs(df$cont_X26.2_GR_MaxLat_dd)
df$cont_X26.3_GR_MinLat_dd <- abs(df$cont_X26.3_GR_MinLat_dd)
df$cont_X26.4_GR_MidRangeLat_dd <- abs(df$cont_X26.4_GR_MidRangeLat_dd)

# log-transform surface and volume traits
# df$cont_X26.1_GR_Area_km2 <- sqrt(df$cont_X26.1_GR_Area_km2)
df$cont_X5.1_AdultBodyMass_g <- log(df$cont_X5.1_AdultBodyMass_g)
df$cont_X5.3_NeonateBodyMass_g <- log(df$cont_X5.3_NeonateBodyMass_g)

# columns we don't want
df <- subset(df, select=-c(
    cont_X27.1_HuPopDen_Min_n.km2,
    cont_X27.2_HuPopDen_Mean_n.km2,
    cont_X27.3_HuPopDen_5p_n.km2,
    cont_X27.4_HuPopDen_Change,
    cont_X17.1_MaxLongevity_m #,
    #disc_X12.2_Terrestriality
  )
)

```

## Imputing missing values

In our data frame there are missing values, although never for more than 50%
of any given column (other columns have already been thrown out). If the amount of missing data
is not too high, there are [valid strategies](http://doi.org/10.1111/2041-210X.12232) for coming
up with approximations of what the missing data might have been given the way the remaining values 
co-vary with each other. There are numerous approaches to this:

## `mice`

Multivariate imputation by chained equations (mice) implements many methods could be used. This is supposed 
to be the best package for doing this without phylogeny:

```{r mice}
library(mice)
mice_result <- mice(df, method = "rf", printFlag = F)
imputed_df <- complete(mice_result)
row.names(imputed_df) <- row.names(df)
```

## `missForest`

Alternatively, missing variables can be imputed using missForest:

```{r missForest}
#library(missForest)
#imputed_df <- missForest(df)[["ximp"]]
```

## Variable selection

First we read in a tree:


```{r phylolm}
library(phylolm)
library(ape)

# read the supertree, randomly resolve
tree <- read.tree(file = "ungulates.nwk")

# drop the tips that had NA records for body mass
tip_df <- imputed_df[tree$tip.label,]

```

Then we do model selection via PGLS:

```{r pgls}
library(nlme)
options(warn=-1)
bm.corr <- corBrownian(phy=tree)
seen <- c()
best <- list()

varsec <- function(df, corr, base_form, seen, best ) {
  nvars <- length(colnames(df)) - 1
  dependents <- colnames(df)[1:nvars]
  result <- list()
  models <- c()
  
  # iterate over all columns
  for ( i in 1:nvars ) {
    
    # skip over already included columns
    if ( ! i %in% seen ) {
      
      # concatenate new form string, parse it
      form_string <- ''
      if ( substr( base_form, start=nchar(base_form), nchar(base_form) ) == '~' ) {
        form_string <- sprintf( "%s%s", base_form, dependents[i] );
      }
      else {
        form_string <- sprintf( "%s+%s", base_form, dependents[i] );
      }
      form <- formula( form_string )
      
      # perform PGLS
      glsr <- gls( form, data = df, correlation = corr, method = "ML" )
      summ <- summary( glsr )
      
      # store results
      result[form_string] <- summ$AIC
      models <- c( models, summ )
    }
  }
  
  # find the lowest AIC, its index, and its formula
  aic <- Inf
  j <- 1
  new_base_form <- ''
  for ( i in 1:length(result) ) {
    if ( result[[i]] < aic ) {
      aic <- result[[i]]
      new_base_form <- names(result)[i]
      j <- i
    }
  }
  print(sprintf("AIC=%f model=%s index=%i", aic, new_base_form, j))
  seen <- c( seen, j )
  best[new_base_form] <- models[[j]]
  
  # recurse or return
  if ( length(seen) < nvars ) {
    varsec( df, corr, new_base_form, seen, best )
  }
  else {
    return(list(AIC = aic, formula = new_base_form))
  }
}

result <- varsec( tip_df, bm.corr, "doms_IsDomesticated~", seen, best )
```

```{r glm}
# the simplified formula with the selected variables
formula = doms_IsDomesticated~cont_X26.7_GR_MidRangeLong_dd+cont_X26.2_GR_MaxLat_dd+cont_X26.5_GR_MaxLong_dd+cont_X26.6_GR_MinLong_dd

res <- phyloglm(formula, tip_df, tree, btol = 30, method = "logistic_IG10", boot = 5)
summary(res)
```