---
title: "Imputing missing values"
author: "Rutger Vos (@rvosa)"
date: "8-12-2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Pre-processing data

We have multiple files with different data types, which we all load separately into data frames:

```{r load_dfs}
df_cont <- read.table("ungulates-continuous.tsv", sep = "\t", header = T, row.names = 1)
df_disc <- read.table("ungulates-discrete.tsv", sep = "\t", header = T, row.names = 1)
df_doms <- read.table("ungulates-domesticated.tsv", sep = "\t", header = T, row.names = 1)
```

Now we want to merge the data frames so that we get one big one, with all the columns as 
in the input frames, and the union of the rows across all frames:

```{r merge}
# SEE: https://adairama.wordpress.com/2017/11/22/how-to-merge-multiple-datasets-in-r-based-on-row-names/
# mimics a recursive merge or full outer join
multimerge <- function (mylist) {
  
  # get the union of the distinct taxon/row names
  unames <- unique(unlist(lapply(mylist, rownames)))
  n <- length(unames)

  # turns input df's into a matrix 'out' with all rows across mylist 
  out <- lapply(mylist, function(df) {
    tmp <- matrix(nr = n, nc = ncol(df), dimnames = list(unames,colnames(df)))
    tmp[rownames(df), ] <- as.matrix(df)
    rm(df); gc()
    return(tmp)
  })
 
  # check if all rownames match unames
  stopifnot( all( sapply(out, function(x) identical(rownames(x), unames)) ) )
  
  # cbind the matrices
  bigout <- do.call(cbind, out)
  
  # set row and column names
  colnames(bigout) <- paste(rep(names(mylist), sapply(mylist, ncol)), unlist(sapply(mylist, colnames)), sep = "_")
  return(bigout)
}

df <- data.frame( multimerge( list( cont=df_cont, disc=df_disc, doms=df_doms ) ) )
rm(df_disc,df_cont,df_doms)
```

And now let's filter this result to remove all the columns that have more than 50% NA:

```{r filter_columns}
df <- df[, colMeans(is.na(df)) <= .5] 
```

## Variable selection

## Imputing missing values

In our data frame there are missing values, although never for more than 50%
of any given column (other columns have already been thrown out). If the amount of missing data
is not too high, there are [valid strategies](http://doi.org/10.1111/2041-210X.12232) for coming
up with approximations of what the missing data might have been given the way the remaining values 
co-vary with each other. There are numerous approaches to this:

## `mice`

Multivariate imputation by chained equations (mice) implements many methods could be used. This is supposed 
to be the best package for doing this without phylogeny:

```{r mice}
library(mice)
mice_result <- mice(df, method = "pmm")
mice_df <- complete(mice_result)
row.names(mice_df) <- row.names(df)
```