---
title: "Ungulate traits and ecogeography"
author: "Rutger Vos (@rvosa)"
date: "7-12-2018"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Trait and geospatial data

We have a [tab-separated table](ungulates-filtered.tsv) containing continuous-valued trait data 
(including latitude and longitude values) from the 
[PanTHERIA](http://esapubs.org/archive/ecol/E090/184/PanTHERIA_1-0_WR05_Aug2008.txt) data base,
filtered on the Ungulates, i.e. the
[Artiodactyla](http://www.departments.bucknell.edu/biology/resources/msw3/browse.asp?id=14200001) and the
[Perissodactyla](http://www.departments.bucknell.edu/biology/resources/msw3/browse.asp?id=14100001).

Here we read that table into a data frame:

```{r readtsv}
# the table has a header, and the 1st column contains the observation names: species
df <- read.table("ungulates.tsv", sep = "\t", header = T, row.names = 1)
```

We can plot the mid points of the latitudinal and longitudinal ranges of the species, on
an `rworldmap`:

```{r plotmap}
library(rworldmap, quietly = T)
newmap <- getMap(resolution = "low")
plot(newmap, asp = 1)
points(df$X26.7_GR_MidRangeLong_dd, df$X26.4_GR_MidRangeLat_dd, col = "red", cex = .6)
```

## Phylogeny

We also have a [phylogenetic tree](ungulates.nwk) in 
[Newick](https://github.com/naturalis/mebioda/blob/master/doc/week1/w1d5/lecture1.md#the-newick--new-hampshire-format)
format. It is the topology of a [supertree of the Mammals](http://doi.org/10.1038/nature05634),
but pruned to the samen taxon set as the trait table. To read that tree, we need to load the `ape` package:

```{r readnwk}
library(ape, quietly = T)
tree <- read.tree(file = "ungulates.nwk")
```

The tree is a `phylo` object, a pseudo-standard in the R community that a number of packages build on. For example,
we can plot the tree against a geological timescale using the `strap` package:

```{r plotphylo}
library(strap, quietly = T)

# set the age of the root to absolute time
tree$root.time <- 88.5
geoscalePhylo(tree, units = "Epoch", cex.age = 1, cex.ts = 0.5, boxes = T, show.tip.label = F)
```

## Combining traits and trees

We can combine the trait data and the tree and plot trait values onto the tree. For example, 
here is the range size of the species, for which we need `phytools`:

```{r plotmass}
library(phytools, quietly = T)

# make a vector of the area column, remove records with NA values
log_area <- log(as.vector(df$X26.1_GR_Area_km2))
names(log_area) <- row.names(df)
log_area <- log_area[!is.na(log_area)]

# drop the tips that had NA records for area
tips_to_prune <- setdiff(row.names(df),names(log_area))
pruned <- drop.tip(tree,tips_to_prune)

# do a continuous trait mapping, of log-transformed mass
contMap(pruned,log_area,standardize=TRUE,length=10,outline=F,lwd=3,ftype="off")
```

## Interaction between traits

We can plot to see how latitude relates to range size:

```{r make_vectors}
# make a vector of absolute latitude, give it names, remove NA values
abs_lat <- abs(df$X26.4_GR_MidRangeLat_dd)
names(abs_lat) <- row.names(df)
abs_lat <- abs_lat[!is.na(abs_lat)]

# keep the intersection for each of the two variables
abs_lat <- abs_lat[intersect(names(abs_lat),names(log_area))]
log_area <- log_area[intersect(names(abs_lat),names(log_area))]
```

The INCORRECT approach would be to just plot the two variables against 
each other even though there is autocorrelation. Here is what that
would look like:

```{r plotwrong}
fitwrong <- lm(log_area ~ abs_lat)
plot(abs_lat,log_area)
abline(fitwrong)
```

A better way to do this is to compute the 
[independent contrasts](https://en.wikipedia.org/wiki/Phylogenetic_comparative_methods#Phylogenetically_independent_contrasts)
and do the modeling on those:

```{r ic}
# drop the tips that had NA records for body mass, randomly resolve
tips_to_prune <- setdiff(row.names(df),names(log_area))
pruned <- multi2di(drop.tip(tree,tips_to_prune))

# compute independent contrasts
abs_lat_contrasts <-pic(abs_lat,pruned)
log_area_contrasts <- pic(log_area,pruned)

# fit linear model
fit <- lm(log_area_contrasts ~ abs_lat_contrasts - 1)
plot(abs_lat_contrasts,log_area_contrasts)
abline(fit)
```

There appears to be correlation:

```{r fitsumm}
summary(fit)
```

## Scaling rules

The result of our analysis shows support for 
[Rapoport's rule](https://en.wikipedia.org/wiki/Rapoport%27s_rule): an ecogeographical 
rule that states that latitudinal ranges of plants and animals are generally smaller at 
lower latitudes than at higher latitudes. In addition to this correlation, our data set
surely shows other correlations, such as trivial scalings:

- Body mass ~ body length
- Longevity ~ body mass
- Neonate body mass ~ gestation length
