---
title: "Trait databases"
author: "Rutger Vos (@rvosa)"
date: "8-12-2017"
output: html_document
---

Trait data web services and semantic integration
------------------------------------------------

Numerous smaller and larger databases exist with ecological or functional
trait data. Many of these databases can be queried directly from within R.
To illustrate this, we are going to collect data for our crop species, i.e.:

```{r vector}
# to pick up as many traits as possible we will consider all putative ancestors
# for each crop (e.g. see the Musa species) and all domesticated "species" 
# (e.g. see the Cucurbita species)
species <- c(
  "Arachis hypogaea",
  "Artocarpus altilis",  
  "Cocos nucifera",
  "Colocasia esculenta",
  "Cucurbita argyrosperma",
  "Cucurbita ficifolia",
  "Cucurbita maxima",
  "Cucurbita moschata",
  "Cucurbita pepo",
  "Dioscorea cayennensis",  
  "Eleusine coracana",  
  "Ensete ventricosum",  
  "Eragrostis tef",
  "Fagopyrum esculentum",
  "Fagopyrum tataricum",
  "Hordeum vulgare",
  "Ipomoea batatas",
  "Mangifera indica",
  "Manihot esculenta",
  "Metroxylon sagu",
  "Musa acuminata",
  "Musa paradisiaca",
  "Musa balbisiana",
  "Oryza sativa",
  "Phoenix dactylifera",
  "Secale cereale",
  "Solanum tuberosum",
  "Sorghum bicolor",
  "Triticum aestivum",  
  "Zea mays"
)
```

For the names of our species, let's check to see if our names are actually recognized, using
the [rOpenSci/taxize package](https://github.com/ropensci/taxize):

```{r taxize}
library(taxize)
accepted_names <- tnrs(query = species, source = "iPlant_TNRS")
accepted_names$acceptedname
```

It turns out that all the names are matched precisely. Let's create a data frame in which we will collect trait data
for our crops from a variety of databases:

```{r frame}
results <- data.frame( row.names = species )
```

Trait databases accessible through GioBo/TR8
--------------------------------------------

The R package [TR8](https://github.com/GioBo/TR8) can fetch trait data
from the following databases:

* [LEDA traitbase](http://www.leda-traitbase.org/LEDAportal/)<sup>*</sup>
* [PLANTS](http://www.bricol.net/)<sup>*</sup>
* [BROT - Mediterranean](http://www.uv.es/jgpausas/brot.htm)
* [Catminat France](http://perso.wanadoo.fr/philippe.julve/catminat.htm)
* [MycoFlor for Central Europe](http://www.esajournals.org/doi/abs/10.1890/12-1700.1)
* [Mycorrhizal intensity database for the USSR](http://esapubs.org/Archive/ecol/E093/059/default.htm)
* [Ecological Flora of California](http://ucjeps.berkeley.edu/efc/)
* [Biolflor Germany](http://www2.ufz.de/biolflor/index.jsp)
* [The Ecological Flora of the British Isles](http://www.ecoflora.co.uk/)
* [Flora Italiana](http://luirig.altervista.org/)
* [Ellenberg values for Italian Flora](http://doi.org/10.7320/FlMedit22.197)

<sup>*</sup>Probably has plant traits globally

Let's fetch data using the `TR8` package and copy it over to the `results` data frame:

```{r tr8}
# installation instructions: https://github.com/GioBo/TR8/blob/master/README.md
library(devtools)
dev_mode(on=T)
library(TR8)

# The dataframe named available_tr8 lists the short codes and descriptions of the traits
# and the respective databases in which they occur. Here we are fetching all of them.
# Alternatively, by providing the gui_config = T argument we can launch a graphical 
# interface that allows us to select traits of interest.
traits_tr8 <- tr8(accepted_names$acceptedname, download_list = available_tr8$short_code)

# Copy the returned traits over to the results data frame
for ( trait in colnames(traits_tr8@results) ) {
  for ( sp in species ) {
    results[ sp, paste( 'tr8', trait, sep = ':' ) ] <- traits_tr8@results[ sp, trait ]
  }
}
```

The bmaitner/RBIEN package
--------------------------

The R package [RBIEN](https://github.com/bmaitner/RBIEN) fetches trait data
from the [botanical information and ecology network](http://bien.nceas.ucsb.edu/bien/),
which we also query for crop plant traits.

```{r bien}
library(BIEN)
library(dplyr)
traits_bien <- BIEN_trait_species(species)
for ( sp in species ) {
  for ( trait in unique(traits_bien$trait_name) ) {
    value <- first( filter( traits_bien, scrubbed_species_binomial == sp & trait_name == trait )$trait_value );
    results[ sp, paste( 'bien', trait, sep = ':' ) ] <- value
  }
}
```

Trait databases accessible through rOpenSci/traits
--------------------------------------------------

The package [traits](https://github.com/ropensci/traits), from the 
[rOpenSci](https://ropensci.org/) initiative, fetches data from these
databases:

* [Encyclopedia of Life Traitbank](http://eol.org/info/516)<sup>*</sup>
* [USDA Plants Database](https://plants.usda.gov/java/)<sup>*</sup>
* [Zanne et al. plant dataset](http://dx.doi.org/10.1038/nature12872)<sup>*</sup>
* [LEDA Traitbase](http://www.uni-oldenburg.de/en/landeco/research/projects/LEDA)
* [BETYdb](http://www.betydb.org)
* [National Center for Biotechnology Information - NCBI](http://www.ncbi.nlm.nih.gov/)
* [Coral Traits Database](http://coraltraits.org/)
* [Amniote life history dataset](https://doi.org/10.1890/15-0846R.1)

<sup>*</sup>Probably has plant traits globally

Let's fetch data from **LEDA** (to compare with TR8), **USDA**, **Zanne et al.**, 
and **EoL**, using the `traits` package.

Starting with **LEDA**, the trait names listed here are all of the traits available in the database. 
Because queries into LEDA require a trait name to look for, we have to hard code them here.

```{r leda}
library(traits)
leda_traits <- c(
  "age_first_flowering", "branching", "buds_seasonality",
  "buds_vertical_dist", "canopy_height", "dispersal_type", "leaf_distribution", "ldmc_geo", "leaf_mass",
  "leaf_size", "morphology_disperal", "growth_form", "life_span", "releasing_height", "seed_longevity", 
  "seed_mass", "seed_number", "seed_shape", "shoot_growth_form", "snp", "ssd", "tv", "clonal_growth_organs"
)

for (trait in leda_traits) {
  leda_record <- leda(trait)
  for ( sp in species ) {
    value <- first(filter(leda_record, sbs_name == sp)[[trait]])
    predicate <- paste( 'leda', trait, sep = ':' )
    if ( is.null(value) ) {
      results[sp, predicate] <- NA
    }
    else {
      results[sp, predicate] <- value
    }
  }
}
```

For the plants traits data dump of the US department of agriculture (**USDA**), the way to invoke it
is by passing the genus and species name as separate arguments. Hence, we will have to split the 
binomial species names on whitespace and then do the function call:


```{r usda}

# The returned results include taxonomic data (i.e. a lot of columns dealing with Linnean classification),
# cultural use, and "actual" traits. We are only interested in the latter:
usda_predicates <- c(
	"Duration", "Growth_Habit", "After_Harvest_Regrowth_Rate", "Bloat", "C_N_Ratio", 
	"Fall_Conspicuous", "Fire_Resistance", "Flower_Color", "Flower_Conspicuous", 
	"Foliage_Color", "Foliage_Porosity_Summer", "Foliage_Porosity_Winter", "Foliage_Texture", 
	"Fruit_Color", "Fruit_Conspicuous", "Growth_Form", "Growth_Rate", 
	"Height_at_Base_Age_Maximum_feet", "Height_Mature_feet", "Leaf_Retention", "Lifespan", 
	"Low_Growing_Grass", "Nitrogen_Fixation", "Resprout_Ability", "Shape_and_Orientation", 
	"Toxicity", "Adapted_to_Coarse_Textured_Soils", "Adapted_to_Medium_Textured_Soils", 
	"Adapted_to_Fine_Textured_Soils", "Anaerobic_Tolerance", "CaCO_3_Tolerance", 
	"Cold_Stratification_Required", "Drought_Tolerance", "Fertility_Requirement", 
	"Fire_Tolerance", "Frost_Free_Days_Minimum", "Hedge_Tolerance", "Moisture_Use", 
	"pH_Minimum", "pH_Maximum", "Precipitation_Minimum", "Precipitation_Maximum", 
	"Root_Depth_Minimum_inches", "Salinity_Tolerance", "Shade_Tolerance", 
	"Temperature_Minimum_F", "Bloom_Period", "Fruit_Seed_Abundance", 
	"Fruit_Seed_Period_Begin", "Fruit_Seed_Period_End", "Fruit_Seed_Persistence", 
	"Propogated_by_Bare_Root", "Propogated_by_Bulbs", "Propogated_by_Container", 
	"Propogated_by_Corms", "Propogated_by_Cuttings", "Propogated_by_Seed", 
	"Propogated_by_Sod", "Propogated_by_Sprigs", "Propogated_by_Tubers", "Seeds_per_Pound", 
	"Seed_Spread_Rate", "Seedling_Vigor", "Small_Grain", "Vegetative_Spread_Rate", 
	"Berry_Nut_Seed_Product", "Palatable_Browse_Animal", "Palatable_Graze_Animal", 
	"Palatable_Human"
)

for ( sp in species ) {
  binomial <- strsplit(sp, " ", fixed = T)
  
  # potentially of interest: `Duration`, `Growth_Habit`, ...
  try(traits_usda <- tr_usda(query = list( genus=binomial[[1]][1], species = binomial[[1]][2] )))
  for ( pr in usda_predicates ) {
    val_usda <- traits_usda[["data"]] %>% 
      select(pr) %>% 
      filter(pr != "") %>% 
      unique %>% 
      first
    results[ sp, paste("usda",pr,sep=":") ] <- val_usda[[1]][1]
  }
}

```

The trait data set of **Zanne et al.** contains only three traits, which we fetch as follows:

```{r zanne}
traits_zanne <- tr_zanne()
for ( sp in species ) {
  
  # handle 'woodiness'
  zw <- traits_zanne$woodiness %>% 
    filter(gs == sp) %>% 
    select(woodiness) %>%
    first
  if (length(zw) > 0) {
    results[sp, "zanne:woodiness" ] <- zw
  }
  
  # handle 'freezing'
  zf <- traits_zanne$freezing %>% 
    filter(species == sp) %>% 
    select(Freeze.tmin.lo) %>% 
    first
  if (length(zf) > 0) {
    results[sp, "zanne:freezing" ] <- zf
  }
  
  # handle 'leaf_phenology'
  zp <- traits_zanne$leaf_phenology %>% 
    filter(Binomial == sp) %>% 
    select(Phenology) %>% 
    first
  if (length(zp) > 0) {
    results[sp, "zanne:leaf_phenology"] <- zp
  }
}

```

The Encyclopedia of Life
------------------------

The [EoL TraitBank](http://eol.org/info/516) can be accessed with the same client packages, but it
is worth paying attention to how the results are represented:

```{r eol}
# EoL's web service sometimes returns an error 500 (server error), so we make
# a function that we can then wrap inside a try(). Function definition:
do_query <- function(eol_taxon_id, sp) {
  eol_traits <- traitbank(eol_taxon_id)
  for ( trait in unique(eol_traits[["graph"]]$predicate) ) {
    results[ sp, trait ] <- first( filter( eol_traits[["graph"]], predicate == trait )$value )
  }
}

# For EoL traitbank data we need to provide the EoL taxon ID as input parameter. Hence,
# we first need to do a TNRS lookup of these, as follows:
sources <- gnr_datasources()
eol_id <- sources[sources$title == "EOL", "id"]
eol_tnrs <- gnr_resolve(species, data_source_ids = c(eol_id), fields = "all")

# We then iterate over the list of species, lookup the EoL taxon ID for that species
# and fetch its traits, which we again copy over to the results frame
for (sp in species) {
  eol_taxon_id <- eol_tnrs[eol_tnrs$matched_name == sp,]$local_id
  try(do_query(eol_taxon_id, sp))
}
```

Once we have all the traits, we will want to remove the columns with excessive amounts of `NA` missing
data. In addition, we will want to remove columns where all species have the same values. The logic
will roughly be:

```{r clean}
# merge rows for the same taxon (e.g. Musa sp.)
#length(na.omit(col)) # should be more than 50% of species, for example
#unique(na.omit(col)) # should be more than 1
```
