---
title: "Trait databases"
author: "Rutger Vos (@rvosa)"
date: "8-12-2017"
output: html_document
---

Collecting trait data for our crops
-----------------------------------

Numerous smaller and larger databases exist with ecological or functional
trait data. Many of these databases can be queried directly from within R.
To illustrate this, we are going to collect data for our crop species, i.e.:

```{r vector}
# to pick up as many traits as possible we will consider all putative ancestors
# for each crop (e.g. see the Musa species) and all domesticated "species" 
# (e.g. see the Cucurbita species)
species <- c(
  "Arachis hypogaea",
  "Artocarpus altilis",  
  "Cocos nucifera",
  "Colocasia esculenta",
  "Cucurbita argyrosperma",
  "Cucurbita ficifolia",
  "Cucurbita maxima",
  "Cucurbita moschata",
  "Cucurbita pepo",
  "Dioscorea cayennensis",  
  "Eleusine coracana",  
  "Ensete ventricosum",  
  "Eragrostis tef",
  "Fagopyrum esculentum",
  "Fagopyrum tataricum",
  "Hordeum vulgare",
  "Ipomoea batatas",
  "Mangifera indica",
  "Manihot esculenta",
  "Metroxylon sagu",
  "Musa acuminata",
  "Musa paradisiaca",
  "Musa balbisiana",
  "Oryza sativa",
  "Phoenix dactylifera",
  "Secale cereale",
  "Solanum tuberosum",
  "Sorghum bicolor",
  "Triticum aestivum",  
  "Zea mays"
)
```

For the names of our species, let's check to see if our names are actually recognized, using
the [rOpenSci/taxize package](https://github.com/ropensci/taxize):

```{r taxize}
library(taxize)
accepted_names <- tnrs(query = species, source = "iPlant_TNRS")
accepted_names$acceptedname
```

(It turns out that all the names are matched precisely.)

Trait databases accessible through GioBo/TR8
--------------------------------------------

The R package [TR8](https://github.com/GioBo/TR8) can fetch trait data
from the following databases:

* [LEDA traitbase](http://www.leda-traitbase.org/LEDAportal/)<sup>*</sup>
* [PLANTS](http://www.bricol.net/)<sup>*</sup>
* [BROT - Mediterranean](http://www.uv.es/jgpausas/brot.htm)
* [Catminat France](http://perso.wanadoo.fr/philippe.julve/catminat.htm)
* [MycoFlor for Central Europe](http://www.esajournals.org/doi/abs/10.1890/12-1700.1)
* [Mycorrhizal intensity database for the USSR](http://esapubs.org/Archive/ecol/E093/059/default.htm)
* [Ecological Flora of California](http://ucjeps.berkeley.edu/efc/)
* [Biolflor Germany](http://www2.ufz.de/biolflor/index.jsp)
* [The Ecological Flora of the British Isles](http://www.ecoflora.co.uk/)
* [Flora Italiana](http://luirig.altervista.org/)
* [Ellenberg values for Italian Flora](http://doi.org/10.7320/FlMedit22.197)

<sup>*</sup>Probably has plant traits globally

Let's fetch data from **LEDA** and **PLANTS** using the `TR8` package:

```{r tr8}
# installation instructions: https://github.com/GioBo/TR8/blob/master/README.md
library(devtools)
dev_mode(on=T)
library(TR8)

# The dataframe named available_tr8 lists the short codes and descriptions of the traits
# and the respective databases in which they occur. Here we are fetching all of them.
# Alternatively, by providing the gui_config = T argument we can launch a graphical 
# interface that allows us to select traits of interest.
traits_tr8 <- tr8(accepted_names$acceptedname, download_list = available_tr8$short_code)
write.csv(traits_tr8@results, file = "traits_tr8.csv")
```

Trait databases accessible through rOpenSci/traits
--------------------------------------------------

The package [traits](https://github.com/ropensci/traits), from the 
[rOpenSci](https://ropensci.org/) initiative, fetches data from these
databases:

* [Encyclopedia of Life Traitbank](http://eol.org/info/516)<sup>*</sup>
* [LEDA Traitbase](http://www.leda-traitbase.org/LEDAportal/)<sup>*</sup>
* [USDA Plants Database](https://plants.usda.gov/java/)<sup>*</sup>
* [Zanne et al. plant dataset](http://dx.doi.org/10.1038/nature12872)<sup>*</sup>
* [BETYdb](http://www.betydb.org)
* [National Center for Biotechnology Information - NCBI](http://www.ncbi.nlm.nih.gov/)
* [Coral Traits Database](http://coraltraits.org/)
* [Amniote life history dataset](https://doi.org/10.1890/15-0846R.1)

<sup>*</sup>Probably has plant traits globally

Let's fetch data from **EoL**, **LEDA** (to compare with TR8), **Euro+Med**, **USDA**
and **Zanne et al.** using the `traits` package:

```{r traits}
library(traits)
library(taxize)

# For EoL traitbank data we need to provide the EoL taxon ID as input parameter. Hence,
# we first need to do a TNRS lookup of these, as follows:
sources <- gnr_datasources()
eol_id <- sources[sources$title == "EOL", "id"]
eol_tnrs <- gnr_resolve(species, data_source_ids = c(eol_id), fields = "all")

# We then iterate over the list of species, lookup the EoL taxon ID for that species
# and fetch its traits
for (sp in species) {
  eol_taxon_id <- unique(eol_tnrs[eol_tnrs$user_supplied_name == sp,]$local_id)
  eol_traits <- traitbank(eol_taxon_id)
}

# For LEDA, see traits::leda(trait = "...")
leda_traits <- c(
  "age_first_flowering", 
  "branching", 
  "buds_seasonality",
  "buds_vertical_dist", 
  "canopy_height",
  "dispersal_type", 
  "leaf_distribution", 
  "ldmc_geo", 
  "leaf_mass",
  "leaf_size", 
  "morphology_disperal", 
  "growth_form", 
  "life_span",
  "releasing_height", 
  "seed_longevity", 
  "seed_mass",
  "seed_number", 
  "seed_shape", 
  "shoot_growth_form",
  "snp", 
  "ssd", 
  "tv", 
  "clonal_growth_organs"
)

leda_out <- list()
for (i in seq_along(leda_traits)) {
  cat(leda_traits[i], sep="\n")
  leda_out[[i]] <- leda(leda_traits[i])
}
sapply(leda_out, NROW)

# For USDA, see traits::tr_usda()
# 1. iterate over species
# 2. per species, fetch data: `tr_usda(query = list(genus = "Oryza", species = "sativa"))`
# 3. potentially of interest: `Duration`, `Growth_Habit`, ...

# For Zanne, see traits::tr_zanne(), e.g.:
# 1. traits_zanne <- tr_zanne()
# 2. traits_zanne[["woodiness"]][ traits_zanne[["woodiness"]]$gs == 'Zea mays', ]
# 3. traits: `woodiness`, `freezing`, `leaf_phenology`
```

The bmaitner/RBIEN package
--------------------------

The R package [RBIEN](https://github.com/bmaitner/RBIEN) fetches trait data
from the [botanical information and ecology network](http://bien.nceas.ucsb.edu/bien/),
which we also query for crop plant traits.

```{r bien}
library(BIEN)
traits_bien <- BIEN_trait_species(species)
```
