---
name: macroevolution
layout: post
title: Some examples to simulate phylogenies and estimate speciation and extinction rates
date: 2015-02-03
author: Tanja Stadler
tags: 
- R
- phylogenetic
- tree shape
---

```{r setup}
library(TreeSim)
library(TreePar)

set.seed(10)
```

### Simulation example - time-dependent rates

First we simulate a tree with a specified number of species:

```{r nspecies}
nspecies <- 100
```

At time 1 in the past, we have a rate shift:

```{r rateshift}
time <- c(0,1)
```

Half of the present day species are sampled (rho[1]=0.5):

```{r sampling}
rho <- c(0.5,1)
```

Speciation rates (between t[i],t[i+1] we have speciation rate lambda[i]):

```{r lambda}
lambda <- c(2,5)
```

Extinction rates (between t[i],t[i+1] we have extinction rate mu[i]):

```{r mu}
mu <- c(1.5,0)
```

Simulation of a tree:

```{r simulate}
tree<-sim.rateshift.taxa( nspecies, 1, lambda, mu, frac=rho, times=time, complete=FALSE )
```

Extracting the speciation times x:

```{r sort}
x<-sort(getx(tree[[1]]),decreasing=TRUE)
```

When estimating the rate shift times t based on branching times x, 
we allow the shift times to be 0.6, 0.8, 1, 1.2, .. ,2.4:

```{r steps}
start <- 0.4
end <- 2
grid <- 0.2
```

We fix rho and estimate time, lambda, mu:

```{r estimate}
res <- bd.shifts.optim(x,c(rho,1),grid,start,end)[[2]]
res
```

res[[2]] tells us about the maximum likelihood estimate given one rate shift:

- log lik = 60.6940763.
- rate shift at time 1.0.
- turnover (extinction/speciation) = 0.68 more recent than 1.0, 
  and = 0.19 more ancestral than 1.0.
- net diversification (speciation-extinction) rate = 0.81 more recent than 1.0, 
  and = 3.66 more ancestral than 1.0.

Values used for simulation:

```{r values}
mu/lambda
lambda-mu
```

Test if 1 shift explain the tree significantly better than 0 shifts:
if test>0.95 then 1 shift is significantly better than 0 shifts at a 5% error

```{r oneshift}
i<-1
test<-pchisq(2*(res[[i]][1]-res[[i+1]][1]),3)
test
```

Test if 2 shifts explain the tree significantly better than 1 shift:

```{r twoshift}
i<-2
test<-pchisq(2*(res[[i]][1]-res[[i+1]][1]),3)
test
```

Plot results:

```{r plot}
bd.shifts.plot(list(res),1,2.1,0,5)

# Plot parameters used for simulation:
lines(c(-max(x),-time[2]),c((lambda-mu)[2],(lambda-mu)[2]),col="red")
lines(c(-time[2],0),c((lambda-mu)[1],(lambda-mu)[1]),col="red")
```

### Simulation example - Test for Diversity-Dependent speciation

Simulate tree:

```{r simdds}
set.seed(1)
tree<-sim.rateshift.taxa( 10, 1, c(2,0.1), c(0,0.05), frac=c(1,1), times=time, complete=FALSE )
x<-sort(getx(tree[[1]]),decreasing=TRUE)
```

(The tree size is too small to make reliable estimations! We just analyse this tree as 
an example for getting answers quickly!)

```{r estimdds}
# Estimate maximum likelihood speciation and extinction rates under a density-dependent speciation model
resDD<-bd.densdep.optim(x,discrete=FALSE,continuous=TRUE)[[2]]
# Slow! Package expoTree on CRAN performs same calculations much faster!! Same method also within Package DDD.

# Estimate maximum likelihood speciation and extinction rates together with the rate shift times
resShifts <- bd.shifts.optim(x,c(rho,1),0.1,0.1,2.1)[[2]][[2]]

resDD
resShifts
```

Best model where AIC smallest:

```{r aic}
AICDD <- 2*3+2*resDD$value
AICShifts <- 2*5+2*resShifts[1]

AICDD
AICShifts
```

### Data example - bird orders

Data analysis in Stadler & Bokma, 2013:
Number of species in each order from Sibley and Monroe (1990)

Load and plot data:

```{r data}
data(bird.orders)
plot(bird.orders)
```

Many species are missing - ie only one per order:

```{r missing}
S <- c(10, 47, 69, 214, 161, 17, 355, 51, 56, 10, 39, 152, 6, 143, 358, 103, 319, 23, 291, 313, 196, 1027, 5712)

# number of species per order:
cbind(bird.orders$tip.label,S)
```

All orders established at 96.43 Ma:

```{r orders}
groupscut<-get.groups(bird.orders,S,96.43) 
x<-branching.times(bird.orders)
```

Transforming molecular timescale into calendar timescale:

```{r transform}
x<-x/0.207407 
```

Allowing one shift in rates:

```{r testoneshift}
resbirds<-bd.shifts.optim(x,sampling=c(1,1),grid=1,start=96,end=135,survival=1,groups=groupscut)[[2]]
i<-1
test<-pchisq(2*(res[[i]][1]-res[[i+1]][1]),3)
test
```

Sampling model is important:

```{r bdshifts}
resbirdsWrongSamp<-bd.shifts.optim(x,sampling=c(length(S)/sum(S),1),grid=1,start=96,end=135,survival=1,groups=0)[[2]]
resbirdsCompleteSamp<-bd.shifts.optim(x,sampling=c(1,1),grid=1,start=96,end=135,survival=1,groups=0)[[2]]
```

Check the difference (especially look at turnover estimates!):

```{r results}
resbirds
resbirdsWrongSamp
resbirdsCompleteSamp
```